Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-11-22 thesis/thesis.tex
thesis: Move and rewrite aims

@@

With Gaia DR2, which contains great quality astrometrical and photometric data for a huge number of stars,
a new opportunity to simplify and optimize this processes is presented.
In this context new approaches are being developed to improve current detection methods and automate the OC characterization.
This work tries to take the best from \emph{Clusterix 2.0. A Virtual Observatory tool to estimate cluster membership probability}
\cite{balaguer2020clusterix} and \emph{Hunting for open clusters in Gaia DR2: 582 new open clusters
in the Galactic disc} \cite{castro2020hunting}, and merge them into a single \emph{machine learning}
tool capable of finding open clusters eficiently without requiring powerful machines to perform the computations.

The first one is a non-parametrized method. It determines empirically the frequency functions asociated to astrometrical variables
without making any assumption about their profiles. However, it heavily does depend on the initial field selection and other
parameters related to frequency functions. \emph{Clusterix 2.0.} is a web app tool inside the Spanish Virtual Observatory
(VO) framework and works together with other VO tool such as:
\emph{TOPCAT} \cite{taylor2005topcat},
\emph{Aladin} \cite{bonnarel2000aladin},
\emph{VOSA} \cite{bayo2008vosa}, among other.

@@

The method developed in this project takes as base the approach made by \citeauthor{castro2020hunting} but rapidly diverges.
In first place, performance has been taken into account since the begining, trying to get a reliable method without compromising the
needs of supercomputers.
To achieve this, instead of blindly analyzing the whole galactic disk a fields selection is made focusing in those regions with known OCs.
Then, these fields are analyzed with an homogeneous methodology based on machine learning algorithms using as sources astrometric and
photometric variables. The goal is the characterization of each individual star in the cluster, discerning wether the star belongs the OC
or not.

Thus, the main goal of this project is not the discovery of new open clusters or even candidates, but the development of a model that,
applied to already knonw open clusters, allows us the characterization of those OCs on an unsupervised way, with a reasonable computing
power and taking into account the same selection of astrometric and photometric variables for every OC.

@@

To measure the reliability of the resulting model comparisons with other VO tools will be made.

The role for the ANN will be validating cluster members based on dynamincs data.

These are te variables to be considered in this work:

To achive this goal, the aim will be divided into the following milestones:

\begin{itemize}
  \item To recover data from Gaia DR2 database. This data will be taken as source for the machine learning model to characterize open clusters
  by grouping stars into clusters.
  \item To research unsupervised clustering algorithms suitable for grouping stars previously recovered.
  \item To select and implement unsupervised clustering algorithms from the previous study.
  \item To apply implemented tools to different datasets from first point to find OCs.
  \item To validate found OCs by determining their accuracy and precission.
\end{itemize}

@@

@@

Due to the large amount of data inside Gaia database, a complete download is not viable neither useful.
In order to squize the size of the dataset to be downloaded, the OpenClust catalogue \cite{dias2002new} (Figure \ref{fig:OpenClustComplete})
has been used to make a preselection of sky regions.

@@

The downloaded dataset covers nearly 114 million stars (42GB of compressed data), that is significantly smaller than the whole Gaia DR2 dataset, which contains
information for around 1,600 million stars, but stils being too large.

@@

Since there is no upper limit to the number of stars inside a cluster, for the sake of simplicity and as a commitment to the project delivery dates
and the available computing power, smaller clusters will be prefered over the greatest ones.

@@

The first one is a container built from a \verb|python:3.8| image that contains the \verb|cdalvaro| package and the \verb|downloader.py| script.
This script is prepared to load the OpenClust catalogue, connect to the Gaia DR2 database, download all stars contained inside each cluster
(with a radius multiplier to increse the area to be downloaded) and save the downloaded data inside the custom database hosted in the second container.

@@

These containers are orchestrated with a \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/docker-compose.yml}{docker-compose.yml}
file which automatilly launches one instance of each container.

@@

This is due to the fact that OC's stars are sourounded by other stars with possibly similar properties. So, setting the number of clusters
to two is too low to separate them properly.

@@

Inside \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/notebooks}{src/notebooks},
there is a notebook called
\href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/notebooks/cluster_characterization.ipynb}{cluster\_characterization.ipynb}
that illustrates how DEC model process a region of stars trying to characterize the OC hidden inside that region.

\section{Results}

@@

\chapter{Discussion}

>>>>

To achieve the general goal, we'll follow the following milestones:

\begin{itemize}
  \item Gather information on the \emph{state of the art}.
  \item Research unsupervised clustering algorithms suitable for grouping stars previously recovered.
  \item Recover data from Gaia DR2 database. This data will be taken as source for the machine learning model to characterize open clusters
  by grouping stars into clusters.
  \item Select and implement unsupervised clustering algorithms from the previous study.
  \item Use those algorithms with different datasets to find OCs.
  \item Look for an independent method to characterize OCs in order to make comparisons.
  \item Validate found OCs with our custom model by making comparisons with the independent method.
\end{itemize}

@@

With Gaia DR2, which contains great quality astrometric and photometric data for a huge number of stars,
a new opportunity to simplify and optimize this processes is presented.
In this context new approaches are being developed to improve current detection methods and automate the OC characterization.
This work tries to take the best from \emph{Clusterix 2.0. A Virtual Observatory tool to estimate cluster membership probability}
\cite{balaguer2020clusterix} and \emph{Hunting for open clusters in Gaia DR2: 582 new open clusters
in the Galactic disc} \cite{castro2020hunting}, and merge them into a single \emph{machine learning}
tool capable of finding open clusters efficiently without requiring powerful machines to perform the computations.

The first one is a non-parametrized method. It determines empirically the frequency functions associated to astrometrical variables
without making any assumption about their profiles. However, it heavily does depend on the initial field selection and other
parameters related to frequency functions. \emph{Clusterix 2.0.} is a web app tool inside the Spanish Virtual Observatory
(VO) framework and works together with other VO tool such as:
\emph{TOPCAT} \cite{taylor2005topcat},
\emph{Aladin} \cite{bonnarel2000aladin},
\emph{VOSA} \cite{bayo2008vosa}, among other.

@@

The method developed in this project takes as base the approach made by \citeauthor{castro2020hunting} but rapidly diverges.
In first place, performance has been taken into account since the beginning, trying to get a reliable method without compromising the
needs of supercomputers.
To achieve this, instead of blindly analyzing the whole galactic disk a fields selection is made focusing in those regions with known OCs.
Then, these fields are analyzed with an homogeneous methodology based on machine learning algorithms using as sources astrometric and
photometric variables. The goal is the characterization of each individual star in the cluster, discerning wether the star belongs the OC
or not.

Thus, the main goal of this project is not the discovery of new open clusters or even candidates, but the development of a model that,
applied to already known open clusters, allows us the characterization of those OCs on an unsupervised way, with a reasonable computing
power and taking into account the same selection of astrometric and photometric variables for every OC.

@@

@@

As mentioned before, we want to characterize clusters looking at their dynamics properties.
So, these are the main features we will take into account for clustering:

@@

Due to the large amount of data inside Gaia database, a complete download is not viable neither useful.
In order to reduce the size of the dataset to be downloaded, the OpenClust catalogue \cite{dias2002new} (Figure \ref{fig:OpenClustComplete})
has been used to make a preselection of sky regions.

@@

The downloaded dataset covers nearly 114 million stars (42GB of compressed data), that is significantly smaller than the whole Gaia DR2 dataset, which contains
information for approximately 1,600 million stars, but it is still being too large.

@@

Since there is no upper limit to the number of stars inside a cluster, for the sake of simplicity and as a commitment to the project delivery dates
and the available computing power, smaller clusters will be preferred over the greatest ones.

@@

The first one is a container built from a \verb|python:3.8| image that contains the \verb|cdalvaro| package and the \verb|downloader.py| script.
This script is prepared to load the OpenClust catalogue, connect to the Gaia DR2 database, download all stars contained inside each cluster
(with a radius multiplier to increase the area to be downloaded) and save the downloaded data inside the custom database hosted in the second container.

@@

These containers are orchestrated with a \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/docker-compose.yml}{docker-compose.yml}
file which automatically launches one instance of each container.

@@

This is due to the fact that OC's stars are surrounded by other stars with possibly similar properties. So, setting the number of clusters
to two is too low to separate them properly.

@@

Inside \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/notebooks}{src/notebooks},
there is a notebook called
\href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/notebooks/cluster_characterization.ipynb}{cluster\_characterization.ipynb}
that illustrates how DEC model processes a region of stars trying to characterize the OC hidden inside that region.

\chapter{Results}

@@

\section{Discussion}
